SPANG: Template Library for Making SPARQL Queries FAIR

### 2020-05-24 paper.tex
v2.0

An increasing number of datasets have been published in the Resource Description Framework (RDF)~\cite{rdf}.
SPARQL is the standardized interface to RDF and contributing effective reuse of distributed RDF data. 
How can we maximize the value of accumulated SPARQL~\cite{sparql} queries?
Reuse of the queries is the key of this problem.
Although eco-systems for developing SPARQL have been developed, writing SPARQL codes often becomes a burden for the Semantic Web users. 
SPARQL queries have been written in many Semantic Web projects, which requires efforts of developers. 
For effective enrichment of the Semantic Web, reuse of such queries are valuable.
% Reusability is the key to maximize the value of accumulated open data and software.
However, due to the limitation of SPARQL specifications, such as the lack of parameterization mechanism, the reuse of those queries are limited. 
% Reusing SPARQL queries are important for efficient development of the Semantic Web.

%The FAIR principle~\cite{fair} could be suitable for addressing this issue.
One of the key principle in recent days of the Semantic Web community is the FAIR (findable, accessible, interoperable and reusable) principle~\cite{fair}.
The Semantic Web is comprised of various specifications, including RDF~\cite{rdf}, OWL~\cite{owl}, and SPARQL~\cite{sparql}. The ontologies written in OWL are accumulated and shared on repositories~\cite{bioportal}.
RDF datasets are also accumulated and hosted in portal web sites~\cite{rdf-portal}.
Not only the ontologies and datasets, but also the queries are important for the Semantic Web.
% As the SPARQL queries are important component of the Semantic Web community,
% they should be shared in a solid principle.
Accumulating and reusing the queries are important.
Notably, the FAIR principle can be applied not only to the RDF data, but also other resource, like software.
% SPARQL is a fundamental technology for querying RDF data contributing an interoperable interface to the Semantic Web. However, is is not designed for reusability. 
% In data management guideline, FAIR principle is well known~\cite{fair}. Findability, Accesibility, Reusablility should be addressed, to make the most of SPARQL queries as resources. 
Applying FAIR principle to SPARQL query should increase the reusability of the queries.

@@

Here, we apply the FAIR principle and extend the SPARQL specifications for more efficient development of the Semantic Web.
%Semantic Web users often write SPARQL queris.
%However, the SPARQL lacks the parameterization mechanism, thus, users write queries for their specific use cases.
%But good SPARQL queries can be reused.
%Actually, writing good query requires much efforts, thus, they should be reused to improve the productivity.
%Thus, considering the reusability of SPARQL queries are crucial to enrich the Semantic Web community.
%How can we improve the reusability of SPARQL?
Following the FAIR principle, we could identify the process to make SPARQL query reusable.
One of the key to improve the reusability of queries is parameterization.
Parameterization make the query more reusable.
Using the template query with parameters, the same query can be applied to many use cases.
% There can be several types of parameterization.
% The SPARQL variables can be considered as parameter.
% In that case, values to the parameters can replace them.
% On the other hands, part of the URI or literals can be parameterized.
% In that case also, values to the parameters can replace them.
% Another mechanism to use template is to give binding to the variables.
%Are SPARQL queries accessible? 
Accessibility to SPARQL queries should also be reconsidered.
afIn most cases, those queries are not accessible. According to the FAIR principle, resources should be accessible by through URI. In some cases, SPARQL queries are stored in repositories. But they do not have stable URIs. How about findability? 
In many Semantic Web projects, 
%SPARQL queries are written and accumulated. However, those SPARQL queries 
%are not findable, due to the lack of export and import SPARQL queries.
SPARQL queries are not exposed to the web in a reusable form, nor indexed by metadata in a findable form.

In this study, we show the step to make SPARQL queries FAIR to maximize the value of written SPARQL queries.
Contributions of this work is demonstrating how to
1) assign meta data to SPARQL queries;
2) parameterize the SPARQL queries;
3) assign global identifiers;
4) construct a server for providing queries.

%%%%%%%%%%%%%%%%%%%%%%%%
\section{SPANG Overview}
%%%%%%%%%%%%%%%%%%%%%%%%

In this study, we review FAIR principles and describe how to apply the principle to SPARQL.
Further, we demonstrate how to use the SPANG template library.
Then, we describe the details of SPANG template specifications.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section{Making SPARQL Queries FAIR}
%%%%%%%%%%%%%%%%%%%%%%%%
Here we review the preliminaries for the FAIR principle
and provide the FAIR principle in the context of SPARQL queries.
The FAIR guiding principle aims to enhance the Findability, Accessibility, Interoperability and Reusability of digital resources.
It shoul be noted that the FAIR principle can be applied not only to the dataset resources but also to other type of resources including software tools.
Not only the RDF dataset, but also the SPARQL queries are important resource to be shared on the Semantic Web platform. Thus, we tried to apply the FAIR principle which have been applied to the RDF datasets, into SPARQL queries.

@@

\subsection{Findability}
The ultimate goal of FAIR is to optimise the reuse of resources. 
To achieve this, metadata should be well-described so that they can be replicated and/or combined in different settings.
In the FAIR principle, making resources findable includes indexing resources by adding metadata to each resources. 
Most of the SPARQL queries are not findable from general users of the Semantic Web. 
The first and most important step toward increased value of the written SPARQL queries is to consider how to make SPARQL queries findable.
The first step is using metadata to find SPARQL queries. 
Metadata should be easy to find for both humans and computers. Machine-readable metadata are essential for automatic discovery of resources, so this is essential component.
% \begin{itemize}
%     \item Assign global identifier
%     \item Rich metadata
%     \item Include the identifier of the data they describe
%     \item Registered and indexed
% \end{itemize}
To make SPARQL queries findable, we propose to add metadata to SPARQL.
The SPARQL metadata includes: title, description, target endpoint, and parameters with its name, default value, and classes.
Indexing of SPARQL queries are conducted and thus users can obtain specific queries of interest through a search API.
We review each item in the following subsections.
In the following subsections, the target of FAIR principle is extended from data to resource.
We consider a SPARQL query as a resource.

\subsection{Accessibility}
To be accessible, resource should be retrievable by identifier, and metadata should be also available even if the resource are no longer available.
%Once the user finds the required data, the user needs to know how can they be accessed, possibly including authentication and authorisation.
% \begin{itemize}
%     \item Retrievable by identifier
%     \item Metadata are accessible
% \end{itemize}
%Some of the SPARQL libraries are accessible from internet. 
%However, we could consider more general scheme for accessibility. 
%In the FAIR principle, accessibility includes minting URI for each resource.
Here we propose to use stable URI for each SPARQL query.
In the case of SPANG library, we assign two URIs for each query. 
One is for retrieving SPARQL query string, thereby users can retrieve the concrete content of the query. 
Another is URI to get results. This URI can include the parameter settings for the specified query.

@@

SPARQL specification is proposed for interoperable use of queries to the Semantic Web. 
It is valuable to use several triplestore implementation through the same programming interface. 
However, standard SPARQL specifications do not include coding rules; capitalization, indentation and newline, etc. 
Lack of those rules result in variation of SPARQL queries, which hampers the reuse of SPARQL queries. 
In other languages, coding rules exist in most cases. 
In the case of SPARQL, even specification document or text books do not follow a specific coding rules. 
In this study, we specify the coding rule used in SPANG template library, and implemented a formatter command to evaluate and reformat the input SPARQL code.

@@

%Reusability of SPARQL is not a new topic. 
In the specification of SPARQL, reusability is not a central topic. 
However, it is an central resource in a Semantic Web, it is valuable to share and reuse SPARQL queires.
Templating is a key for reusability. 
SPARQL specifications do not includes templating mechanism, but without templating, each SPARQL queries are not applicable to other use cases. 
We implemented templating mechanism using a general freamework, mustache.
Various activity is conducted to construct libraries of SPARQL. 
It is considered as activities to increase the reusability of SPARQL queries. However, no standard to increase the reusability is proposed. 
%If it is a component of Semantic Web comprised of RDF/OWL/SPARQL, it is reasonable to represent it in RDF framework. 
%We propose a way to represent SPARQL in the RDF framework.

\subsection{FAIRification of SPARQL}
The process of making SPARQL FAIR ("FAIRification") can be described in multiple steps~\cite{fairification}.
Most of the requirements for findability and accessibility can be achieved at the metadata level. 
Interoperability and reusability requires more efforts at the SPARQL code level.

@@

\end{verbatim}
\end{scriptsize}
\caption{Example query in SPANG library}
\label{fig:example-sparql}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%
\section{SPANG Library and Use Cases}
%%%%%%%%%%%%%%%%%%%%%%%%%

The core of the SPANG specification is the SPARQL metadata and templating mechanism for increasing the reusability of the SPARQL queries.

An example query in the SPANG library is shown in Figure~\ref{fig:example-sparql}.
The DisGeNet endpoint~\cite{disgenet} is specified.

@@

As a use case, we developed a SPANG API to search for the set of SPARQL queries.

@@

We propose the parapeterization mechanism of SPARQL query. It is crucial to reuse the SPARQL query for us to improve the productivity of the programming experiance on the Semantic Web platform.

Using SPANG API, users can execute SPARQL query over the Web, with the given parameters.

\subsection{SPANG Library API}
According to the metadata and parameterization mechanism, we can construct a controlled set of SPARQL library which is reusable in various settings.

SPANG library is accesible through SPANG Web API.

The following command returns the list of libraries.

The results is shown in Figure~\ref{fig:libs-api-out} as JSON format.

The following command returns the list of query in a library.

The results is shown in Figure~\ref{fig:lib-api-out} as JSON format.

\subsection{SPANG Query API}

The following command returns the result of the query, with the given paramter.

@@

Figure~\ref{fig:spang_disease_gene_query} shows a query for DiesGeNet SPARQL endpoint to get disase gene relationship. 
Figure~\ref{fig:spang_disease_gene_result} shows the result of the query. Thus, users can get the results only by clicking on the web pages, browsing the SPARQL template on the SPANG web site.

The following command searches for queries with keywords.

\subsection{SPANG Server}
SPANG is originally developed for "SARQLing" the RDF dataset from command line. Here, we re-implemented the SPANG using Node.js and now the functionality is available through Web browser.

@@

Figure~\ref{fig:spang_lib} is the list of SPARQL templates for DisGeNet SPARQL endpoint.
A short name for the SPARQL template, descriptin of the query, and example parameters are shown.

>>>>

An increasing number of datasets have been published in the Resource Description Framework (RDF)~\cite{rdf}.
SPARQL is the standardized interface to RDF and contributing effective reuse of distributed RDF data. 
How can we maximize the value of accumulated SPARQL~\cite{sparql} queries?
Reuse of the queries is the key of this problem.
Although eco-systems for developing SPARQL have been developed, writing SPARQL codes often becomes a burden for the Semantic Web developers and users. 
SPARQL queries have been written in many Semantic Web projects, which requires efforts. 
For effective construction of the Semantic Web applications, reuse of such queries are valuable.
% Reusability is the key to maximize the value of accumulated open data and software.
However, due to the limitation of SPARQL specifications, such as the lack of parameterization mechanism, the reuse of those queries are limited. 
% Reusing SPARQL queries are important for efficient development of the Semantic Web.

%The FAIR principle~\cite{fair} could be suitable for addressing this issue.
One of the key principle in recent days of the Semantic Web community is the FAIR (findable, accessible, interoperable and reusable) principle~\cite{fair}.
The ultimate goal of FAIR is to optimise the reuse of resources. 
The Semantic Web is comprised of various specifications, including RDF~\cite{rdf}, OWL~\cite{owl}, and SPARQL~\cite{sparql}. The ontologies written in OWL are accumulated and shared on repositories~\cite{bioportal}.
RDF datasets are also accumulated and hosted in portal web sites~\cite{rdf-portal}.
Not only the ontologies and datasets, but also the queries are important for the Semantic Web.
% As the SPARQL queries are important component of the Semantic Web community,
% they should be shared in a solid principle.
Accumulating and reusing the queries are important.
Notably, the FAIR principle can be applied not only to the RDF data, but also other resource, like software.
% SPARQL is a fundamental technology for querying RDF data contributing an interoperable interface to the Semantic Web. However, is is not designed for reusability. 
% In data management guideline, FAIR principle is well known~\cite{fair}. Findability, Accesibility, Reusablility should be addressed, to make the most of SPARQL queries as resources. 
Applying FAIR principle to SPARQL query should increase the reusability of the queries.

@@

Here, we apply the FAIR principle and extend the SPARQL specifications for more efficient development of the Semantic Web.
%Semantic Web users often write SPARQL queris.
%However, the SPARQL lacks the parameterization mechanism, thus, users write queries for their specific use cases.
%But good SPARQL queries can be reused.
%Actually, writing good query requires much efforts, thus, they should be reused to improve the productivity.
%Thus, considering the reusability of SPARQL queries are crucial to enrich the Semantic Web community.
%How can we improve the reusability of SPARQL?
Following the FAIR principle, we could identify the process to make SPARQL query reusable.
One of the key to improve the reusability of queries is parameterization.
Parameterization make the query more reusable.
Using the template query with parameters, the same query can be applied to many use cases.
% There can be several types of parameterization.
% The SPARQL variables can be considered as parameter.
% In that case, values to the parameters can replace them.
% On the other hands, part of the URI or literals can be parameterized.
% In that case also, values to the parameters can replace them.
% Another mechanism to use template is to give binding to the variables.
%Are SPARQL queries accessible? 
Accessibility to SPARQL queries should also be reconsidered.
In most cases, those queries are not accessible. According to the FAIR principle, resources should be accessible by through URI. In some cases, SPARQL queries are stored in repositories. But they do not have stable URIs. How about findability? 
In many Semantic Web projects, 
%SPARQL queries are written and accumulated. However, those SPARQL queries 
%are not findable, due to the lack of export and import SPARQL queries.
SPARQL queries are not exposed to the web in a reusable form, nor indexed by metadata in a findable form.

In this study, we show the steps to make SPARQL queries FAIR to maximize the value of written SPARQL queries:
1) assign meta data to SPARQL queries;
2) parameterize the SPARQL queries;
3) assign global identifiers;
4) construct a server for providing queries.
And we demonstrate the steps to construct the SPANG template library, which is reusable on the Semantic Web platform.

%%%%%%%%%%%%%%%%%%%%%%%%
\section{SPANG Project Overview}
%%%%%%%%%%%%%%%%%%%%%%%%

@@

In the following sections, we review FAIR principles and describe how to apply the principle to SPARQL.
Further, we demonstrate how to use the SPANG template library.
Then, we describe the details of SPANG template specifications.

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section{Making SPARQL Queries FAIR}
%%%%%%%%%%%%%%%%%%%%%%%%

The FAIR guiding principle aims to enhance the Findability, Accessibility, Interoperability and Reusability of digital resources.
It shoul be noted that the FAIR principle can be applied not only to the dataset resources but also to other type of resources including software tools.
Not only the RDF dataset, but also the SPARQL queries are important resource to be shared on the Semantic Web platform. Thus, we tried to apply the FAIR principle which have been applied to the RDF datasets, into SPARQL queries.
In the following subsections, we review the preliminaries for the FAIR principle
and provide the FAIR principle in the context of SPARQL queries.
We review each item in the following subsections.
In the following subsections, the target of FAIR principle is extended from data to resource.
We consider a SPARQL query as a resource.

@@

\subsection{Findability}
The ultimate goal of FAIR is to optimise the reuse of resources. 
To achieve this, metadata should be well-described so that they can be replicated and/or combined in different settings.
In the FAIR principle, making resources findable includes indexing resources by adding metadata to each resources. 
Most of the SPARQL queries are not findable from general users of the Semantic Web. 
The first and most important step toward increased value of the written SPARQL queries is to consider how to make SPARQL queries findable.
The first step is using metadata to find SPARQL queries. 
Metadata should be easy to find for both humans and computers. Machine-readable metadata are essential for automatic discovery of resources, so this is essential component.
% \begin{itemize}
%     \item Assign global identifier
%     \item Rich metadata
%     \item Include the identifier of the data they describe
%     \item Registered and indexed
% \end{itemize}
To make SPARQL queries findable, we propose to add metadata to SPARQL.
The SPARQL metadata includes: title, description, target endpoint, and parameters with its name, default value, and classes.
Indexing of SPARQL queries are conducted and thus users can obtain specific queries of interest through application programming interfaces (APIs).

\subsection{Accessibility}
To be accessible, resource should be retrievable by identifier.
%Once the user finds the required data, the user needs to know how can they be accessed, possibly including authentication and authorisation.
% \begin{itemize}
%     \item Retrievable by identifier
%     \item Metadata are accessible
% \end{itemize}
%Some of the SPARQL libraries are accessible from internet. 
%However, we could consider more general scheme for accessibility. 
%In the FAIR principle, accessibility includes minting URI for each resource.
Here we propose to use stable URI for each SPARQL query.
In the case of SPANG library, we assign two URIs for each query. 
One is for retrieving SPARQL query string, thereby users can retrieve the concrete content of the query. 
Another is URI to get results. This URI can include the parameter settings for the specified query.
Also, according to the FAIR principle, metadata of each resource should be also available even if the resource are no longer available. 
We assume that API can provide the metadata for each SPARQL query even if the query is no longer available.

@@

SPARQL specification is proposed for interoperable use of queries to the Semantic Web. 
It is valuable to use several triplestore implementation through the same programming interface. 
However, standard SPARQL specifications do not include coding rules; capitalization, indentation and newline, etc. 
Lack of those rules result in variation of SPARQL queries, which hampers the efficiency in collaborative development of SPARQL queries. 
In other languages, coding rules exist in most cases. 
In the case of SPARQL, even specification document or text books do not follow a specific coding rules. 
Thus, we specify the coding rule used in SPANG template library, and implemented a formatter command to evaluate and reformat the input SPARQL code.

@@

%Reusability of SPARQL is not a new topic. 
The ultimate goal of FAIR is to optimise the reuse of resources. 
However, in the specification of SPARQL, reusability is not a central topic. 
Since it is an central resource in a Semantic Web, it is valuable to share and reuse SPARQL queries.
Templating is a key for reusability. 
SPARQL specifications do not includes templating mechanism, but without templating, each SPARQL queries are not applicable to other use cases. 
We implemented templating mechanism using a general freamework, mustache.
Various activity is conducted to construct libraries of SPARQL. 
It is considered as activities to increase the reusability of SPARQL queries. However, no standard to increase the reusability is proposed. 
%If it is a component of Semantic Web comprised of RDF/OWL/SPARQL, it is reasonable to represent it in RDF framework. 
%We propose a way to represent SPARQL in the RDF framework.

\subsection{FAIRification of SPARQL}
The process of making resource FAIR ("FAIRification") can be described in multiple steps~\cite{fairification}.
Most of the requirements for findability and accessibility can be achieved at the metadata level. 
Interoperability and reusability requires more efforts at the resource level.
The same strategy can be applied to the SPARQL queries.
The following section describes how we constructed the SPANG library which follows the FAIR principle.

@@

\end{verbatim}
\end{scriptsize}
\caption{Example query in the SPANG library}
\label{fig:example-sparql}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%
\section{SPANG Library and Use Cases}
%%%%%%%%%%%%%%%%%%%%%%%%%
In the previous section we review the FAIR principle in the context of SPARQL queries.
In this section, we describe how we constructed the SPANG library with some use cases.
The use case shown in this section includes the query to the DisGeNet endpoint~\cite{disgenet}, which is shown in Figure~\ref{fig:example-sparql}.
This query retrieve protein IDs involved in a specific disease.
The core of the SPANG library specification is the SPARQL metadata and templating mechanism for increasing the reusability of the SPARQL queries.
The query example includes metadata at the beginning of the query as comment lines.
The query also includes a parameter for a disease identifier.
According to the metadata and parameterization mechanism, we can construct a controlled set of SPARQL library which is reusable in various settings.

@@

We developed a SPANG library API to search for the set of SPARQL queries.
SPANG library is accessible through SPANG Web API.
The following command returns the list of libraries.

The results is shown in Figure~\ref{fig:libs-api-out} as JSON format.
The following command returns the list of query in a library.

The results is shown in Figure~\ref{fig:lib-api-out} as JSON format.
The following command searches for queries with keywords.

@@

We propose the parameterization mechanism of SPARQL query. It is crucial to reuse the SPARQL query for us to improve the productivity of the programming experience on the Semantic Web platform.

Using SPANG API, users can execute SPARQL query over the Web, with the given parameters.
The following command returns the result of the query, with the given parameter.

@@

Figure~\ref{fig:spang_disease_gene_query} shows a query for DisGeNET SPARQL endpoint to get disease gene relationship. 
Figure~\ref{fig:spang_disease_gene_result} shows the result of the query. Thus, users can get the results only by clicking on the web pages, browsing the SPARQL template on the SPANG web site.

\subsection{SPANG Server}
SPANG is originally developed as a SPARQL client to query the RDF dataset from command line. Here, we re-implemented the SPANG using Node.js and now the functionality is available through Web browsers.

@@

Figure~\ref{fig:spang_lib} is the list of SPARQL templates for DisGeNET SPARQL endpoint.
A short name for the SPARQL template, description of the query, and example parameters are shown.
